Tropical Mathematics for the Lambda Calculus

### 2023-01-12 main.tex
commit
Push serale
Non so se dopo mangiato avrò le forze per scrivere qualcosina in più.
La sezione 3 va ricontrollata, aggiustata ecc, ma ora ci sono tutti gli elementi.

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{tikz-cd}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{amsthm}
\usetikzlibrary{cd}
\usepackage{stmaryrd}

@@

\title{Lambda Calculus through Tropical Mathematics}

@@

\section{Two Quantitative Approaches to the $\lambda$-calculus}
\input{section2}

@@

\section{Analytic Results}
\input{section4}%

@@

\section{Quasi-metric Spaces and $\Lawv$-Modules}\label{sec:QMod}
\input{section6}%

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{tikz-cd}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{amsthm}
\usetikzlibrary{cd}
\usepackage{mathabx}
\usepackage{cmll}
\usepackage{stmaryrd}

@@

\title{Tropical Mathematics for the Lambda Calculus}

@@

\section{A Bridge Between Two Quantitative Approaches to the $\lambda$-calculus}
\input{section2}

@@

\section{Tropical Laurent series}
\input{section4}%

@@

\section{Generalized Metric Spaces and Modules over the Lawvere Quantale}\label{sec:QMod}
\input{section6}%

### 2023-01-12 section3B.tex

It is known ([reference??]) that the maps $X\mapsto \C M_{\leq n}(X)$ ($n\in\BB N$) lift to functors $!_n:\LREL\to\LREL$.
The sequence $(!_n)_{n\in\N}$ can be then turned into a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, satisfying the adjunction:
$\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
Therefore, $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$. 
Concretely, a type $A::= o \mid !_{n}A \multimap A$ (i tipi meglio dirli in Section II) is in interpreted as $\model o=\B 1$ and 
$\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.

@@

It is instructive at this point to see what its CCC-structure looks like in our tropical world.
In particular, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t:\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
\[
 \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1*\cdots *\mu_n}
 \set{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}}.
\]
The exponential object $X\multimap Y$ is $!X+Y$ (where $+$ is the disjoint union).
Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation morphism is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) + X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.

@@

\begin{example}
 The function $f:\Lawv\to\Lawv$, $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series: it is of shape $f=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not descrete. Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
\end{example}

>>>>

It is known {\color{red}(([reference??])} that the maps $X\mapsto \C M_{\leq n}(X)$ ($n\in\BB N$) lift to functors $!_n:\LREL\to\LREL$.
The sequence $(!_n)_{n\in\N}$ can be then turned into a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, satisfying the adjunction:
$\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
Therefore, $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$. 
Concretely, a type %$A::= o \mid !_{n}A \multimap A$
{\color{red}(i tipi meglio dirli in Section II)} is in interpreted as $\model o=\B 1$ and 
$\model{!_{n}A\multimap B}:= \C M_{\leq n}(\model A) \times \model B$.

@@

It is instructive at this point to see what its CCC-structure looks like in our tropical world.
In particular, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
\[
 \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1*\cdots *\mu_n}
 \set{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}}.
\]
The exponential object $X\multimap Y$ is $!X+Y$ (where $+$ is the disjoint union).
Remember that in $\Lawv$ the neutral element for addition is $\infty$ and the neutral for multiplication is $0$, so for instance the evaluation map is the matrix $\RM{eval}^{X,Y}\in\Lawv^{!((X\multimap Y) + X)\times Y}\simeq\Lawv^{(!!X\times !Y\times !X)\times Y}$ given by $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=0$ if $\rho_1=[\mu]$ and $\rho_2=[b]$, and $\RM{eval}^{X,Y}_{\rho_1\oplus\rho_2\oplus\mu,b}:=\infty$ otherwise.

@@

\begin{example}\label{ex:famous_ex}
 The function $f:\Lawv\to\Lawv$, $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series: it is of shape $f=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not descrete. Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
\end{example}

### 2023-01-12 section4.tex
Sezioni riorganizzate
Ora tutte le sezioni del sorgente dovrebbero corrispondere alle sezioni del pdf
commit

@@

- concavity

- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.

- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
\begin{itemize}

\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally

\item singular points: the famous example

\item tropical roots

\end{itemize}

@@

- characterization of the functional metric using derivatives

>>>>

In this section we study the tropical Laurent series $\Lawv^X\to \Lawv ^Y$ from the viewpoint of analysis.

Let us start by recalling Example~\ref{ex:famous_ex}.
By plotting its graph {\color{red}vogliamo plottarlo?}, we see first of all that the function is non-decreasing and concave.
It is easy to see that this is actually always the case:

\begin{proposition}
 Any tropical Laurent series $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
\end{proposition}

The $f$ of Example~\ref{ex:famous_ex} is continuous on $\BB R_{\geq0}=\Lawv-\set{\infty}$ (w.r.t.\ the usual norm of real numbers).
By considering the usual norm $\norm{x}_\infty:=\sup\limits_{a\in X} \absv{x_a}$ on $\Lawv^X$, we could generalise this property by dropping the case of $x$ having some $0$ coordinate:

\begin{theorem}\label{thm:cont}
 Any tropical Laurent series $f:\Lawv^X\to\Lawv$ is continuous on $\BB R_{>0}$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
\end{theorem}
\begin{proof}
 The result follows after proving that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
\end{proof}

Not only the $f$ of Example~\ref{ex:famous_ex} is continuous, but it is also locally Lipschitz on $\BB R_{>0}$.
Actually, we can prove the following:

\begin{theorem}
 Let $f:\Lawv\to\Lawv$ a tropical Laurent series with matrix $\widehat f:\N\to\Lawv$.
 For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N$ s.t.:
 \begin{enumerate}
  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv$;
  \item If $f(x_0)=\infty$ for some $x_0<\infty$, then $\C F_\epsilon=\emptyset$;
  \item On all $[\epsilon,\infty]$, $f$ coincides with the tropical \emph{polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\widehat f(n)}$.
 \end{enumerate}
 In particular, $f$ is locally Lipschitz on all $\BB R_{>0}$.
\end{theorem}
\begin{proof}
 We can let $\C F_\epsilon:=\set{n\in\N \mid
 \widehat f(n)<\infty \textit{ and } \widehat f(m)> \widehat f(m)+\epsilon \textit{ for all } m<n}$.
\end{proof}

Remark that, in coherence with the previous result, in Example~\ref{ex:famous_ex} $f(x)$ is indeed a $\min$ for all $x>0$.
At $x=0$ we have $f(x=0)=\inf\limits_{n\in\N} \frac{1}{2^n}=0$ which is \emph{not} a min.
Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
This phenomenon is reminiscent of [Example 7, PCoh]
%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
, which actually motivated our first investigations.

We could generalise the locally Lipschitz property as follows:
\begin{theorem}
 Let $f:\Lawv^X\to\Lawv$.
 If $f$ is non-decreasing, concave and continuous, then it is locally Lipschitz.
\end{theorem}
\begin{proof}
 Refinement of the arguments used for Theorem~\ref{thm:cont}.
 {\color{red}Che ci scriviamo?}
\end{proof}

Highlight compositional reasoning based on Lipschitz.\\

@@

{\color{red}
- extend this result to functions $\Lawv^{X}\to \Lawv^{Y}$ with $X,Y$ finite (as this will be useful in section V) Quale ?}

@@

- characterization of the functional metric using derivatives\\

Let us end this section by mentioning another point of view on tropical Laurent series.
$\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete normed cone.
The normed cone structure induces an order on it, called its \emph{cone order}, by setting:
$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
This order makes it a Scott-continuous dcpo.
Furthermore we have:

\begin{proposition}
  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}$, w.r.t.\ the cone orders on the domain and codomain.
\end{proposition}
